---
title: "Financial Viability Analysis"
author: "Vicky Wang (viwa)"
date: "2024-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# install these packages if you don't have them already installed
library(ggplot2)
library(dplyr)
library(plyr)
library(reshape2)
library(knitr)
library(binom) 
library(purrr)
library(forcats)
library(tidyr)
library(stringr)
library(scales)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dx11 <- read.csv("9DX11.csv") # read in diagnoses index
op11 <- read.csv("9OP11.csv") # read in operation index
crit <- read.csv("CRIT.csv") # read in critical care ICU patient info
dispostn <- read.csv("DISPOSTN.csv") # read in disposition
drg <- read.csv("DRG.csv") # read in diagnostic-related groups
ome_dx <- read.csv("OME_DX.csv") # read in omega diagnostic data
ome_px <- read.csv("OME_PX.csv") # read in omega procedure data
omega <- read.csv("OMEGA.csv") # patient encounter data
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# this code chunk is designed to clean the dx11 data set so it can be usable in R
# filters out any rows with nothing in it-- this happens when exporting the accb into .csv
dx11 <- dx11 %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0) # cleans data of empty rows

dx11 <- dx11 %>%
  mutate(ALL = str_trim(ALL)) %>%  # Trim leading and trailing white spaces
  mutate(ALL = str_remove(ALL, "^\\|")) %>%  # Remove leading pipe if present
  mutate(ALL = str_remove(ALL, "\\|$")) %>%  # Remove trailing pipe if present
  separate(ALL, 
           into = c("DIAG", "DIAG_DESC", "DIAG_C", "DIAG_5"), #new column names
           sep = " \\| ",   # Assume there's a space before and after each pipe
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("DIAG"), ~str_trim(.))) %>% # Trim white space from new columns
  select(DIAG, DIAG_DESC, DIAG_C, DIAG_5) %>% #removes extra info we don't want
  mutate(DIAG = as.integer(DIAG)) %>% #sets the columns into their relevant type
  mutate(DIAG_DESC = as.character(DIAG_DESC)) %>% #sets into string
  mutate(DIAG_C = as.double(DIAG_C)) %>% #sets into number
  mutate(DIAG_5 = as.double(DIAG_5)) #sets into number

dx11 <- dx11[-1, ] #removes the double column names situation
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# this code chunk is designed to clean op11 data set so it is usable in R
op11 <- op11 %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0)

op11 <- op11 %>%
  mutate(ALL = str_trim(ALL)) %>% 
  mutate(ALL = str_remove(ALL, "^\\|")) %>%
  mutate(ALL = str_remove(ALL, "\\|$")) %>%
  separate(ALL, 
           into = c("OP", "OP_DESC", "OP_C"), 
           sep = " \\| ",
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("OP"), ~str_trim(.))) %>%
  select(OP, OP_DESC, OP_C) %>%
  mutate(OP = as.double(OP)) %>%
  mutate(OP_DESC = as.character(OP_DESC)) %>%
  mutate(OP_C = as.double(OP_C))

op11 <- op11[-1, ]
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# this code chunk is designed to clean crit data set so it is usable in R
crit <- crit %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0)

crit <- crit %>%
  mutate(ALL = str_trim(ALL)) %>% 
  mutate(ALL = str_remove(ALL, "^\\|")) %>%
  mutate(ALL = str_remove(ALL, "\\|$")) %>%
  separate(ALL, 
           into = c("PAT_NO", "PAT_NO_TAG", "ADM_DATE", "DIS_DATE", "TOT_CCUS", "OCCURRENCE", "ICU", "ICU_LOS"), 
           sep = " \\| ",
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("PAT_NO"), ~str_trim(.))) %>%
  select(PAT_NO, PAT_NO_TAG, ADM_DATE, DIS_DATE, TOT_CCUS, OCCURRENCE, ICU, ICU_LOS)

crit <- crit[-1, ]
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# this code chunk is designed to clean dispostn data set so it is usable in R
dispostn <- dispostn %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0)

dispostn <- dispostn %>%
  mutate(ALL = str_trim(ALL)) %>% 
  mutate(ALL = str_remove(ALL, "^\\|")) %>%
  mutate(ALL = str_remove(ALL, "\\|$")) %>%
  separate(ALL, 
           into = c("DISPOSTN", "DESCRIPT"), 
           sep = " \\| ",
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("DISPOSTN"), ~str_trim(.))) %>%
  select(DISPOSTN, DESCRIPT)

dispostn <- dispostn[-1, ]
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# this code chunk is designed to clean drg data set so it is usable in R
drg <- drg %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0)

drg <- drg %>%
  mutate(ALL = str_trim(ALL)) %>% 
  mutate(ALL = str_remove(ALL, "^\\|")) %>%
  mutate(ALL = str_remove(ALL, "\\|$")) %>%
  separate(ALL, 
           into = c("DRG", "MAJ_DX_CAT", "PARTITION", "DESCRIPTOR", "WEIGHT", "ARITH_LOS", "MEAN_LOS", "CUTOFF"), 
           sep = " \\| ",
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("DRG"), ~str_trim(.))) %>%
  select(DRG, MAJ_DX_CAT, PARTITION, DESCRIPTOR, WEIGHT, ARITH_LOS, MEAN_LOS, CUTOFF)

drg <- drg[-1, ]
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# this code chunk is designed to clean ome_dx data set so it is usable in R
ome_dx <- data.frame(ALL = ome_dx)

ome_dx <- ome_dx[which(nchar(gsub("\\s", "", ome_dx$ALL)) > 0), , drop = FALSE]

# We use the read.table function to separate the ALL column into multiple columns
ome_dx <- read.table(text = ome_dx$ALL, sep = "\t", header = TRUE, fill = TRUE, stringsAsFactors = FALSE)
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# cleans ome_px so it is usable for R
ome_px <- data.frame(ALL = ome_px)

ome_px <- ome_px[which(nchar(gsub("\\s", "", ome_px$ALL)) > 0), , drop = FALSE]

# We use the read.table function to separate the ALL column into multiple columns
ome_px <- read.table(text = ome_px$ALL, sep = "\t", header = TRUE, fill = TRUE, stringsAsFactors = FALSE)
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
#removes empty rows in omega
omega <- omega %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0)
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# removes whitespace and separates ALL column into respective table columns. also adjusts variable type from defaulted "string"
omega <- omega %>%
  mutate(ALL = str_trim(ALL)) %>% 
  mutate(ALL = str_remove(ALL, "^\\|")) %>%
  mutate(ALL = str_remove(ALL, "\\|$")) %>%
  separate(ALL, 
           into = c("PAT_NO", "ADM_DATE", "DIS_DATE", "ADM_HR", "ADM_TYPE", "ADM_SOURCE", "ADM_SVC", "ADM_DIAG_9", "ADM_LISTA9", "AGE", "AGE_DAYS", "BIRTH_WGHT", "PCONSULTS", "DCONSULTS", "CCSTAY", "PRINCPL_DX", "AUTOPSY", "CORONER", "DIED_WI_48", "DIED_IN_OR", "DIED_PSTOP", "DOA", "ER_DEATH", "DISPOSTN", "DIS_SVC", "DISTANCE", "PPX_DOP", "WEIGHT", "DRG", "EXT_REV", "LISTA_9", "LISTB_9", "LOS", "LOS_B_SURG", "MORPH_NEO", "MORPH_BEH", "PRINCPL_OP", "PRIM_PAY", "SEC_PAY", "MAJOR_PAY", "PERC90", "PHYS_A_KEY", "PHYS_A_SPC", "PHYS_KEY", "PHYS_SPC", "PREPRC_TST", "RACE", "ELOS_UNTR", "FAM_REF", "OTH_PROVDR", "RELIGION", "SEX", "PPX_SURG", "TEACH_SVC", "TISSUE", "ZIP", "DIS_HOUR", "LOS_HRS", "DSCHG_RESD", "ADDL_SVC", "MOTHER_MRN", "OP_ER_PHYS", "COMP_PAY", "ORGAN_DON", "ORGAN_TYPE", "BILLING_NO", "OUTPT_COST", "ADMT_RESD", "SUPV_RESD", "PAT_TYPE", "TOT_CHARGE", "TOT_PAY", "TOT_COST", "TOT_DIRECT"), 
           sep = " \\| ",
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("PAT_NO"), ~str_trim(.))) %>%
  mutate(PAT_NO = as.integer(PAT_NO)) %>%
  mutate(ADM_HR = as.integer(ADM_HR)) %>%
  mutate(ADM_SOURCE = as.integer(ADM_SOURCE)) %>%
  mutate(ADM_DATE = str_trim(ADM_DATE)) %>%
  mutate(DIS_DATE = str_trim(DIS_DATE)) %>%
  mutate(ADM_TYPE = str_trim(ADM_TYPE)) %>%
  mutate(ADM_SVC = as.integer(ADM_SVC)) %>%
  mutate(ADM_DIAG_9 = as.double(ADM_DIAG_9)) %>%
  mutate(ADM_LISTA9 = as.integer(ADM_LISTA9)) %>%
  mutate(AGE = as.integer(AGE)) %>%
  mutate(AGE_DAYS = as.integer(AGE_DAYS)) %>%
  mutate(BIRTH_WGHT = as.double(BIRTH_WGHT)) %>%
  mutate(PCONSULTS = as.integer(PCONSULTS)) %>%
  mutate(DCONSULTS = as.integer(DCONSULTS)) %>%
  mutate(CCSTAY = as.integer(CCSTAY)) %>%
  mutate(PRINCPL_DX = as.double(PRINCPL_DX)) %>%
  mutate(AUTOPSY = as.integer(AUTOPSY)) %>%
  mutate(CORONER = as.integer(CORONER)) %>%
  mutate(DIED_WI_48 = as.integer(DIED_WI_48)) %>%
  mutate(DIED_IN_OR = as.integer(DIED_IN_OR)) %>%
  mutate(DIED_PSTOP = as.integer(DIED_PSTOP)) %>%
  mutate(ER_DEATH = as.integer(ER_DEATH)) %>%
  mutate(DISPOSTN = as.integer(DISPOSTN)) %>%
  mutate(DIS_SVC = as.integer(DIS_SVC)) %>%
  mutate(DISTANCE = as.integer(DISTANCE)) %>%
  mutate(WEIGHT = as.double(WEIGHT)) %>%
  mutate(DRG = as.integer(DRG)) %>%
  mutate(TOT_CHARGE = as.double(TOT_CHARGE)) %>%
  mutate(TOT_PAY = as.double(TOT_PAY)) %>%
  mutate(TOT_COST = as.double(TOT_PAY)) %>%
  mutate(TOT_DIRECT = as.double(TOT_DIRECT))
omega <- omega[-1, ]

omega <- omega %>% distinct()
```


``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# cleans and reads service so it is usable in r
service <- read.csv("SERVICE.csv")
service <- service %>% 
  rowwise() %>% 
  mutate(count_separators = str_count(ALL, fixed("|"))) %>% 
  ungroup() %>% 
  filter(count_separators != 0)

service <- service %>%
  mutate(ALL = str_trim(ALL)) %>% 
  mutate(ALL = str_remove(ALL, "^\\|")) %>%
  mutate(ALL = str_remove(ALL, "\\|$")) %>%
  separate(ALL, 
           into = c("SERVICE", "DESCRIPT"), 
           sep = " \\| ",
           extra = "merge",
           fill = "right") %>% 
  mutate(across(starts_with("SERVICE"), ~str_trim(.)))
service <- service[-1, ]
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# writing new files that look way better for analysis in r
write.csv(dx11, file = "9dx11_new.csv", row.names = FALSE)
write.csv(op11, file = "9op11_new.csv", row.names = FALSE)
write.csv(crit, file = "crit_new.csv", row.names = FALSE)
write.csv(dispostn, file = "dispostn_new.csv", row.names = FALSE)
write.csv(drg, file = "drg_new.csv", row.names = FALSE)
write.csv(ome_dx, file = "ome_dx_new.csv", row.names = FALSE)
write.csv(ome_px, file = "ome_px_new.csv", row.names = FALSE)
write.csv(omega, file = "omega_new.csv", row.names = FALSE)
write.csv(service, file = "service_new.csv", row.names = FALSE)
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
mdc_list <- read.csv("MDC_List.csv") # read in major diagnostic groups file that has been exported to csv form
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
drg <- drg %>%
  mutate(MAJ_DX_CAT = as.integer(MAJ_DX_CAT)) #makes maj_dx_cat an integer

mdc_list <- mdc_list %>%
  mutate(MDC = as.integer(MDC)) #makes MDC an integer
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
add_mdc <- drg %>%
  left_join(mdc_list, by = c("MAJ_DX_CAT" = "MDC")) %>%
  mutate(DRG = as.integer(DRG)) #joins the mdc_list to drg using MAJ_DX_CAT


find_mdc <- omega %>%
  left_join(add_mdc, by = c("DRG"))

filter_services <- find_mdc %>%
  filter(DESCRIPT == "Cardiology" | DESCRIPT == "Orthopedics") #joins mdc to omega using DRG and also filters to services of interest

```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
write.csv(find_mdc, file = "omega-card-ortho_filtered.csv", row.names = FALSE)
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
profit <- find_mdc %>%
  filter(DESCRIPT != "" & trimws(DESCRIPT) != "") %>%
  mutate(SERVICE_TYPE = case_when(
    DESCRIPT == "Cardiology" ~ "Cardiology",
    DESCRIPT == "Orthopedic" ~ "Orthopedics",
    TRUE ~ "Other"
  ))

library(dplyr)

```

**Profitability of patients from Cardiac and Orthopedic services**
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
#USING GROSS MARGIN CALCULATIONS TO ASSESS PROFITABILITY: TOTAL PAID - ALL DIRECT COSTS / DIRECT COST
# Nguyễn, H. N. (2010). Marketing Metrics The Definitive Guide to Measuring Marketing Performance (2nd Edition)-Pearson Prentice Hall (2010). Prentice Hall. https://www.academia.edu/81941821/Marketing_Metrics_The_Definitive_Guide_to_Measuring_Marketing_Performance_2nd_Edition_Pearson_Prentice_Hall_2010_

# Calculate the averages by service type
service_type_averages <- find_mdc %>%
  filter(DESCRIPT != "" & trimws(DESCRIPT) != "") %>%
  mutate(SERVICE_TYPE = case_when(
    DESCRIPT == "Cardiology" ~ "Cardiology",
    DESCRIPT == "Orthopedic" ~ "Orthopedics",
    TRUE ~ "Other"))

summary_gross_margin <- service_type_averages %>%
  dplyr::group_by(SERVICE_TYPE) %>%
  dplyr::summarise(
    AVG_PAY = mean(TOT_PAY),
    AVG_DIRECT = mean(TOT_DIRECT)
  ) %>%
  mutate(AVG_GROSS_MARGIN = (AVG_PAY - AVG_DIRECT) / AVG_DIRECT) %>%
  mutate(percentage = AVG_GROSS_MARGIN*100)
```

Table of Average Gross Margin, Calculated on Total Pay and Direct Cost Values for each Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
summary_gross_margin %>%
  select(SERVICE_TYPE, percentage) %>%
  kable(digits = 2, col.names = c("Type of Service", "Average Gross Margin (%)"))

```

Average Gross Margin by Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
summary_gross_margin %>%
  ggplot(mapping = aes(x = SERVICE_TYPE, y = percentage, fill = SERVICE_TYPE)) +
  geom_bar(stat = "identity") + ggtitle("Gross Margin by Service Type") + xlab("Type of Service") + ylab("Average Gross Margin (%)") + scale_fill_discrete(name = "Service Type")
```


**LOS Outliers**
Plot of Non-Outliers vs Outliers in Length of Stay by their Service Types
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
profit <- profit %>%
  mutate(LOS = as.integer(LOS))


IQR_values <- profit %>%
  filter(!is.na(LOS)) %>% # filter out missing LOS values
  summarize(
    Q1 = quantile(LOS, 0.25, na.rm = TRUE),
    Q3 = quantile(LOS, 0.75, na.rm = TRUE),
    IQR = IQR(LOS, na.rm = TRUE)
  ) %>%
  mutate(
    lower_bound = Q1 - 1.5 * IQR,
    upper_bound = Q3 + 1.5 * IQR
  ) %>%
  {list(lower = .$lower_bound, upper = .$upper_bound)}

# Identify outliers and create a new column 'Outlier'
profit <- profit %>%
  mutate(
    Outlier = case_when(
      LOS < IQR_values$lower | LOS > IQR_values$upper ~ "Outlier",
      TRUE ~ "Not Outlier"
    )
  )

# Create your ggplot
p <- ggplot(profit, aes(x = SERVICE_TYPE, y = LOS, color = Outlier)) +
  geom_point(trim = FALSE) +  # Create the violin plot
  geom_jitter(width = 0.2) +  # Add points with a slight horizontal "jitter" to enhance visibility
  labs(x = "Service Type", y = "Length of Stay (LOS)") + ggtitle("Length of Stay vs Service Type (Blue Denotes Outlier)") +
  theme_minimal() +
  theme(legend.position = "none") +  # Remove the legend if color by group is not necessary
  scale_color_brewer(palette = "Set1") # Use a color palette for distinction

# Print the plot
print(p)

```

Table of Proportion of Outliers in Length of Stay for each Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
outliers_los <- profit %>%
  dplyr::group_by(SERVICE_TYPE) %>%
  dplyr::mutate(num_values = n()) %>%
  filter(Outlier == "Outlier") %>%
  dplyr::mutate(num_outliers = n(), .groups = "drop") %>% # Use n() to count rows in the filtered group
  dplyr::summarise(prop_outliers = num_outliers/num_values) %>%
  distinct(SERVICE_TYPE, prop_outliers) %>%
  kable(digits = 2, col.names = c("Service Type", "Proportion of Outliers per Type"))

outliers_los

```

**Payer mix**

1. Pie chart of the overall summary of primary payors, regardless of major diagnostic category group.
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
tot_num_row <- nrow(profit)

find_pie1 <- profit %>%
  dplyr::group_by(PRIM_PAY) %>%
  dplyr::summarise(N_PAYOR = n()) %>%
  dplyr::mutate(PERCENT_PAYER = N_PAYOR/tot_num_row)
```



``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
  #group= this will be prim_pay LETTERS[1:5],
  #this will be percent_payor value=c(13,7,9,21,2)

# Compute the position of labels
pie_create1 <- find_pie1 %>% 
  arrange(desc(PRIM_PAY)) %>%
  mutate(prop = PERCENT_PAYER / sum(find_pie1$PERCENT_PAYER) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )

# Basic piechart

# Create a named vector with a color for each level of PRIM_PAY
custom_colors <- setNames(c("#F8766D", "#B79F00", "#00BA38", "#00BFC4", "#619CFF", "#F564E3"), levels(profit$PRIM_PAY))

# Create a named vector with a new label for each level of PRIM_PAY
custom_labels <- setNames(c("Blue Cross", "Commercial", "Medicaid", "Medicare", "Private", "Work Comp"), levels(profit$PRIM_PAY))

# Ensure this vector length matches the number of levels
ggplot(pie_create1, aes(x="", y=prop, fill=PRIM_PAY)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  theme(legend.position="right") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Primary Payor (All)") 
```
Most patient encounters are paid for by the Blue Cross, Commercial payors, or Medicare. Medicaid represents a smaller proportion, and Private/Work Comp insurers are least frequent.


2. Summary of Primary Payors for Cardiology Services
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
tot_num_row <- nrow(profit)

find_pie2 <- profit %>%
  filter(SERVICE_TYPE == "Cardiology") %>%
  dplyr::group_by(PRIM_PAY) %>%
  dplyr::summarise(N_PAYOR = n()) %>%
  dplyr::mutate(PERCENT_PAYER = N_PAYOR/tot_num_row)
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
  #group= this will be prim_pay LETTERS[1:5],
  #this will be percent_payor value=c(13,7,9,21,2)

# Compute the position of labels
pie_create2 <- find_pie2 %>% 
  arrange(desc(PRIM_PAY)) %>%
  mutate(prop = PERCENT_PAYER / sum(find_pie2$PERCENT_PAYER) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )

# Ensure this vector length matches the number of levels
ggplot(pie_create2, aes(x="", y=prop, fill=PRIM_PAY)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  theme(legend.position="right") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Primary Payor (Cardiology)") 
```
Most (50%) cardiology-service patient encounters are paid using Medicare.

3. Summary of Primary Payors for Orthopedics Services.
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
tot_num_row <- nrow(profit)

find_pie3 <- profit %>%
  filter(SERVICE_TYPE == "Orthopedics") %>%
  dplyr::group_by(PRIM_PAY) %>%
  dplyr::summarise(N_PAYOR = n()) %>%
  dplyr::mutate(PERCENT_PAYER = N_PAYOR/tot_num_row)
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
  #group= this will be prim_pay LETTERS[1:5],
  #this will be percent_payor value=c(13,7,9,21,2)

# Compute the position of labels
pie_create3 <- find_pie3 %>% 
  arrange(desc(PRIM_PAY)) %>%
  mutate(prop = PERCENT_PAYER / sum(find_pie3$PERCENT_PAYER) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )

# Ensure this vector length matches the number of levels
ggplot(pie_create3, aes(x="", y=prop, fill=PRIM_PAY)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  theme(legend.position="right") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Primary Payor (Orthopedic)") 
```

Most orthopedic-service patient encounters are paid using Medicare.


4. Summary of Primary Payors for all Other Services.
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
tot_num_row <- nrow(profit)

find_pie4 <- profit %>%
  filter(SERVICE_TYPE == "Other") %>%
  dplyr::group_by(PRIM_PAY) %>%
  dplyr::summarise(N_PAYOR = n()) %>%
  dplyr::mutate(PERCENT_PAYER = N_PAYOR/tot_num_row)
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
  #group= this will be prim_pay LETTERS[1:5],
pie_create4 <- find_pie4 %>% 
  arrange(desc(PRIM_PAY)) %>%
  mutate(prop = PERCENT_PAYER / sum(find_pie4$PERCENT_PAYER) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )

# Ensure this vector length matches the number of levels
ggplot(pie_create4, aes(x="", y=prop, fill=PRIM_PAY)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  theme(legend.position="right") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Primary Payor (Other)") 
```
Most non-cardiology or non-orthopedic services are paid for using commercial insurers.

**Rates of emergent vs. elective care**

Bar Graph of Admission Type, Separated by Service Type 
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
e_care <- profit %>%
  dplyr::filter(ADM_TYPE %in% c("ELECTIVE", "EMERGENT")) %>%
  dplyr::group_by(SERVICE_TYPE, ADM_TYPE) %>%
  dplyr::summarise(num_type = n(), .groups = "keep") %>%
  dplyr::group_by(SERVICE_TYPE) %>%
  dplyr::mutate(total_service_type = sum(num_type)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(prop_type = num_type / total_service_type) %>%
  dplyr::select(SERVICE_TYPE, ADM_TYPE, prop_type)

e_care %>%
    kable(digits = 3, col.names = c("Service Type", "Type of Care", "Proportion per Type"))
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# Create a named vector with a color for each level of PRIM_PAY
custom_colors <- setNames(c("#F8766D", "#619CFF"), levels(e_care$ADM_TYPE))

# Create a named vector with a new label for each level of PRIM_PAY
custom_labels <- setNames(c("Elective", "Emergent"), levels(e_care$ADM_TYPE))

ggplot(e_care, aes(x = SERVICE_TYPE, y = prop_type, fill = ADM_TYPE)) +
  geom_bar(stat = "identity", position = "stack") +
  ggtitle("Proportion of Elective vs Emergent Care by Service Type") +
  xlab("Service Type") +
  ylab("Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Admission Type") 

```

**Probable future demand given Omega's patient demographics for Cardiac and Orthopedic services** 
(this is less of a coding/query question, and more of discussing what you can an infer from demographics and trends)

Number of Patients, Separated by Age Group and Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
age_demo <- profit %>%
  select(PAT_NO, ADM_DATE, AGE, AGE_DAYS, RACE, SEX, SERVICE_TYPE, PRIM_PAY, LOS, TOT_PAY, TOT_CHARGE, TOT_DIRECT, TOT_COST) %>%
  filter(AGE != "" & trimws(AGE) != "") %>%
  dplyr::mutate(AGE_GROUP = case_when(
    AGE_DAYS > 0 ~ "< 2 years",
    AGE >= 2 & AGE < 18 ~ "2-17 years",
    AGE >= 18 & AGE <= 34 ~ "18-34 years",
    AGE >= 35 & AGE <= 44 ~ "35-44 years",
    AGE >= 45 & AGE <= 54 ~ "45-54 years",
    AGE >= 55 & AGE <= 64 ~ "55-64 years",
    AGE >= 65 ~ "65+ years",
  )) %>%
  dplyr::group_by(SERVICE_TYPE, AGE_GROUP) %>%
  dplyr::summarise(num_people = n()) 

custom_colors <- setNames(c("#F8766D", "#619CFF", "#00BA38"), levels(age_demo$SERVICE_TYPE))

# Create a named vector with a new label for each level of PRIM_PAY
custom_labels <- setNames(c("Cardiology", "Orthopedics", "Other"), levels(age_demo$SERVICE_TYPE))

age_demo %>%
  ggplot(mapping = aes(x = AGE_GROUP, y = num_people, fill = SERVICE_TYPE)) + geom_bar(stat = "identity") +
  ggtitle("Number of People per Age Group, by Service Type") + xlab("Age Group") + ylab("Number of Patients") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Service Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Number of Patients, Separated by Race and Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
race_demo <- profit %>%
  filter(RACE != "UNKNOWN") %>%
  select(PAT_NO, ADM_DATE, AGE, AGE_DAYS, RACE, SEX, SERVICE_TYPE, PRIM_PAY, LOS, TOT_PAY, TOT_CHARGE, TOT_DIRECT, TOT_COST) %>%
  dplyr::group_by(SERVICE_TYPE, RACE) %>%
  dplyr::summarise(num_people = n())

custom_colors <- setNames(c("#F8766D", "#619CFF", "#00BA38"), levels(race_demo$SERVICE_TYPE))

# Create a named vector with a new label for each level of PRIM_PAY
custom_labels <- setNames(c("Cardiology", "Orthopedics", "Other"), levels(race_demo$SERVICE_TYPE))

race_demo %>%
  ggplot(mapping = aes(x = RACE, y = num_people, fill = SERVICE_TYPE)) + geom_bar(stat = "identity") +
  ggtitle("Number of People by Race, Separated by Service Type") + xlab("Race") + ylab("Number of Patients") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Service Type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Number of Patients, Separated by Sex and Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
sex_demo <- profit %>%
  filter(SEX != "UNKNOWN") %>%
  select(PAT_NO, ADM_DATE, AGE, AGE_DAYS, RACE, SEX, SERVICE_TYPE, PRIM_PAY, LOS, TOT_PAY, TOT_CHARGE, TOT_DIRECT, TOT_COST) %>%
  dplyr::group_by(SERVICE_TYPE, SEX) %>%
  dplyr::summarise(num_people = n())

custom_colors <- setNames(c("#F8766D", "#619CFF", "#00BA38"), levels(sex_demo$SERVICE_TYPE))

# Create a named vector with a new label for each level of PRIM_PAY
custom_labels <- setNames(c("Cardiology", "Orthopedics", "Other"), levels(sex_demo$SERVICE_TYPE))

sex_demo %>%
  ggplot(mapping = aes(x = SEX, y = num_people, fill = SERVICE_TYPE)) + geom_bar(stat = "identity") +
  ggtitle("Number of People by Sex, Separated by Service Type") + xlab("Sex") + ylab("Number of Patients") +
  scale_fill_manual(values = custom_colors, labels = custom_labels, name = "Service Type") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Number of Patients, Separated by RACE, AGE GROUP, SEX, and Service Type
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
all_demo <- profit %>%
  filter(RACE != "UNKNOWN" & SEX != "UNKNOWN" & !is.na(AGE)) %>%
  dplyr::mutate(AGE_GROUP = case_when(
    AGE_DAYS > 0 ~ "< 2 years",
    AGE >= 2 & AGE < 18 ~ "2-17 years",
    AGE >= 18 & AGE <= 34 ~ "18-34 years",
    AGE >= 35 & AGE <= 44 ~ "35-44 years",
    AGE >= 45 & AGE <= 54 ~ "45-54 years",
    AGE >= 55 & AGE <= 64 ~ "55-64 years",
    AGE >= 65 ~ "65+ years",
  )) %>%
  select(PAT_NO, ADM_DATE, AGE_GROUP, RACE, SEX, SERVICE_TYPE, PRIM_PAY, LOS, TOT_PAY, TOT_CHARGE, TOT_DIRECT, TOT_COST) %>%
  dplyr::group_by(SERVICE_TYPE, RACE, AGE_GROUP, SEX) %>%
  dplyr::summarise(num_people = n())

custom_colors <- setNames(c("#F8766D", "#619CFF", "#00BA38"), levels(all_demo$SERVICE_TYPE))

# Create a named vector with a new label for each level of PRIM_PAY
custom_labels <- setNames(c("Cardiology", "Orthopedics", "Other"), levels(all_demo$SERVICE_TYPE))

ggplot(all_demo, aes(x = AGE_GROUP, y = num_people, fill = RACE)) +
  geom_bar(stat = "identity", position = position_stack()) +
  facet_grid(SERVICE_TYPE ~ SEX, scales = "free_x", space = "free_x") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(angle = 0)) +
  xlab("Age Group") +
  ylab("Number of People") +
  ggtitle("Demographic Distribution by Service Type and Race")
```

``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
library(lubridate)  # For easy date manipulation
```

Number of Patients of each Service Type over Aug 1999 - Dec 2000
``` {r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, warning=FALSE, dpi=80}
# Convert ADM_DATE to Date, extract year and month, and calculate monthly patient counts
# Group by month and service type, then calculate the count
over_time_patients <- profit %>%
  mutate(ADM_DATE = as.Date(ADM_DATE, format = "%m/%d/%Y")) %>%
  mutate(ADM_MONTH = floor_date(ADM_DATE, "month")) %>%  # Creates a monthly date column
  dplyr::group_by(ADM_MONTH, SERVICE_TYPE) %>%
  dplyr::summarise(num_patients = n(), .groups = 'drop') %>%
  arrange(ADM_MONTH, SERVICE_TYPE)

# Plot the total number of patients by month and service type
ggplot(over_time_patients, aes(x = ADM_MONTH, y = num_patients, group = SERVICE_TYPE, color = SERVICE_TYPE)) +
  geom_line() +
  geom_point() +
  labs(title = "Number of Patients Over Time by Service Type",
       x = "Month",
       y = "Number of Patients",
       color = "Service Type") +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```